Vagueness \& Rationality

### 2018-08-27 springer/current/e1.tex
did a bit more on e1, still needs stats numbers imported from Rmds

In Experiment 1, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more. An example stimulus is given in Figure \ref{Experiment1and2examplestimulus}.

Examples of crisp and vague versions of the numerical and verbal instructions are given next: the examples assume the array (6,15,24) and reference to the smaller number of dots, such that 6 was classified as the expected response; 15 as the borderline response; and 24 as the incorrect response. In the {\em vague numerical} condition we used \emph{Choose a square with about 10 dots}. None of the squares contained 10 dots. 10 is slightly closer to 6 than to 15, justifying 6 as the best response and 15 as the borderline response. In the {\em vague verbal} condition we used \emph{Choose a square with few dots}. In the {\em crisp numerical} condition we used \emph{Choose the square with 6 dots}, and one square always did contain the number mentioned. For {\em crisp verbal}, we used \emph{Choose the square with the fewest dots}.

@@

\begin{description}
	\item [Hypothesis 1] There exists a main effect RT advantage for vagueness.
	\item [Hypothesis 2] An RT advantage for vagueness both in \emph{numeric} and in \emph{verbal} instructions.
	\item [Hypothesis 3] No large main effect of instruction format (i.e., numeric versus verbal) since we hypothesised that vagueness rather than instruction format drove the effect in Experiment 1.
	\item [Hypothesis 4] On the basis of our experiment in \citet{green2013utility}, we expect faster responses for stimuli with more discriminable arrays.(i.e., an effect of item).
	\item [Hypothesis 5] Participants should make more borderline case choices for vague than crisp instructions.
\end{description}

\subsection{Experimental Method (Experiment 1)}

% Participants were aged between 18 and 45 with a median age of 28. 
We manipulated as independent variables vagueness and instruction format (i.e., numeric versus verbal instructions), yielding four conditions, \emph{vague numeric}; \emph{vague verbal}; \emph{crisp numeric}; \emph{crisp verbal}.
We measured two dependent variables: response time; and the probability of a participant choosing the borderline case. On each trial, first the referring expression that constituted the instruction for that trial was displayed. 
Participants then pressed a key to indicate that they had read the instruction. After 1000 ms, the arrays were presented, while preserving the text of the referring expression. The response time dependent variable was measured from the presentation of the arrays, until the keypress indicating the participant's choice, which was also recorded. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as `correct' or `incorrect', but instead as `borderline response', or `not borderline response', and we did not want to draw participants' attention to this distinction explicitly. We simply recorded whether the participant chose the best referent, the borderline case or the poorest referent, and how long it took the participant to respond.

\subsection{Results (Experiment 1)}

Means for response times and proportion of borderline responses are given in Fig. (\ref{resultsC-exp-1}). Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. A linear mixed model was constructed for the (logged) response times, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:

@@

\begin{description}
	\item [Test of Hypothesis 1] The main effect of \emph{vagueness} was to slow responses down, in contrast with Experiment 1, and offering evidence against Hypothesis 1 (vague: 2668 ms; crisp: 2450 ms; a difference of 218 ms; $\beta=.06$, $se=.01$, $t=4.6$, $p<0.0001$). 
	\item [Test of Hypothesis 2] In focussed comparisons, vagueness was significantly disadvantageous in the \emph{numeric} conditions ($\beta=0.09$, $se=0.02$, $t=4.5$, $p<0.0001$), and non-significantly disadvantageous in the \emph{verbal} conditions ($\beta=0.02$, $se=0.02$, $t=1.5$, $p=0.1550$), offering partial evidence against Hypothesis 2. The disadvantage for vagueness was greater in the numerical than in the verbal conditions, leading to a significant overall interaction effect between vagueness and instruction format ($\beta=-0.13$, $se=0.02$, $t=-6.6$, $p=0.0169$).
	\item [Test of Hypothesis 3] There was a significant effect of \emph{instruction format} with numerical conditions attracting longer responses than the verbal conditions: consistent with Experiment 1, but suggesting that instruction format rather than vagueness drove the effect we observed in \citet{green2013utility} (numeric: 3284 ms; verbal 1866 ms; a difference of 1418 ms; ($\beta=-0.36$, $se=0.07$, $t=-5.1$, $p<0.0001$).
	\item [Test of Hypothesis 4] There was a significant main effect of item (i.e., of which triple of numbers of dots were used in the stimulus): ($\beta=0.12$, $se=0.02$, $t=7.1$, $p<0.0001$). This effect seems likely to be due to the very fast responses for stimuli using arrangements of the smallest numbers of dots (6,15,24), which also had the largest difference in ratio of smallest number to largest number in the stimulus, suggesting that stimuli using these numbers of dots may have been particularly discriminable for participants.
\end{description}

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultsC-exp-1}. Participants chose the borderline square on $16.6\%$ of trials overall.

\begin{description}
	\item [Test of Hypothesis 5] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{description}

@@

Experiment 1 tested to see whether when borderline cases are present, vague instructions would speed responses as they did in our experiment in \citet{green2013utility} when there were no borderline squares.

We actually found a small (but statistically significant) \emph{disadvantage} of vague instructions: vague instructions slowed people down by 112 ms on average. We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average, such that the disadvantage of numerical format overwhelmed the contribution of vagueness. The \emph{verbal vague} condition was still responded to faster than the \emph{numerical crisp} condition, so the pattern from \citet{green2013utility} was reproduced, but in the light of the evidence from Experiment 1, in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like an advantage of verbal instruction format.

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: To see this, consider the task of selecting the dot array that contains ``few dots'': to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains ``16 dots'' seems to require the participant to estimate, and then {\em match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm''.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. In Experiment 2 we restricted all the instructions to \emph{numeric} quantifiers while factorially manipulating vagueness and selection task. In Experiment 3 we ensured that all instructions used \emph{verbal} quantifiers, while also factorially manipulating vagueness and selection task. This allowed us to distinguish between the predictions of the selection task account and the instruction format account.

>>>>

In Experiment 1, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more, and these flanking arrays appeared equally often on the left and right of the central array.

Table \ref{instructionsC-exp-1} gives the full set of stimuli and associated instructions. 
%
The way in which borderline responses were construed is as follows, using as an example the array (6:15:24) and instructions that identified the smaller flanking array (6). 6 was classified as the expected response. 15 was classified as the borderline response. 24 was classified as the extreme response.

\begin{itemize}
\item In the vague numerical condition the instruction was "Choose a square with about 10 dots" -- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response. 
\item In the vague verbal condition we used "Choose a square with few dots". We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\item In the crisp numerical condition we used "Choose the square with 6 dots". The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\item For crisp verbal, we used "Choose the square with the fewest dots". We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., "Choose a square with about 10 dots"). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure \ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. 
%
In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as "correct" or "incorrect", but instead as "expected response"; "borderline response"; and "extreme response", and we did not want to draw participants' attention to this distinction explicitly. Which choice the participant made was recorded for analysis.

@@

\begin{description}
	\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions should result in faster responses than crisp instructions; and this pattern should hold when the model is restricted to numeric-only data and when it is restricted to verbal-only data.
	\item [Hypothesis 2 (Numeric/Verbal RT)] There should be no real difference between responses to Numeric instructions and Verbal instructions (based on our interpretation of the experiment in \citet{green2013utility}, where we thought that vague instructions alone were driving the advantage for instructions that were both vague, and also in verbal format).
	\item [Hypothesis 3 (Item RT)] Responses should take longer as the number of dots in the display grows larger (i.e., as the levels of Item increase).
	\item [Hypothesis 4 (Response Type)] Vague instructions should lead to more borderline responses than crisp instructions.
\end{description}

\subsection{Results (Experiment 1)} 
\subsubsection{Full model}
\marginnote{needs numbers for the stats}

Means for response times and proportion of borderline responses are given in Fig. (\ref{resultsC-exp-1}). Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. A linear mixed model was constructed for the (logged) response times, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and per-participant intercepts and slopes for sum-coded vagueness, instruction format, (and their interaction) as random effects. We found:

@@

\begin{description}
\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions actually led to significantly slower responses than crisp instructions, against Hypothesis 1: $\beta=$, $se=$, $t=$, $p$. When the model was restricted to numeric-only instructions Vague instructions still led to significantly slower responses than crisp instructions $\beta=$, $se=$, $t=$, $p$. When the model was restricted to verbal-only instructions Vague instructions tended to slow responses, but not significantly: $\beta=$, $se=$, $t=$, $p$.
\item [Hypothesis 2 (Numeric/Verbal RT)] There was actually a significant difference between numeric and verbal instructions, with numeric instructions leading to longer responses than verbal instructions, against Hypothesis 2: $\beta=$, $se=$, $t=$, $p$.
\item [Hypothesis 3 (Item RT)] Responses took longer as the levels of Item increased, supporting Hypothesis 3: $\beta=$, $se=$, $t=$, $p$.
\end{description}

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant, as well as per-participant intercepts as random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultsC-exp-1}. Participants chose the borderline square on $16.6\%$ of trials overall.

\begin{description}
\item [Hypothesis 4 (Response Type)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{description}

\subsubsection{Restricted model}

However, given that the plot shows that responses to 6:15:24 in the "crisp numeric" instructions condition were extremely fast relative to the "vague numeric" instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference.

A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this results in the effects tabled below.

With less data available, the model formula had to be simplified in order to converge -- specifically the following terms were dropped: per-subject slopes for the Vagueness by Item interaction, and per-subject slopes for the effect of Item.

On the basis of the restricted model:

\begin{description}
\item [Hypothesis 1 (Crisp/Vague RT)] After dropping the 6:15:24 level, there is still a disadvantage for Vagueness, but it is not significant (p=.24) $\beta=$, $se=$, $t=$, $p$. When the model was restricted to numeric-only instructions the disadvantage for vagueness was very small and not significant $\beta=$, $se=$, $t=$, $p$. When the model was restricted to verbal-only instructions the disadvantage for vagueness was small and not significant $\beta=$, $se=$, $t=$, $p$.
\item [Hypothesis 2 (Numeric/Verbal RT)] There is still a significant disadvantage for Numeric instruction format (p<.001).
\item [Hypothesis 3 (Item RT)] There is still a disadvantage for increasing the number of dots but it is not significant (p=.323). 
%\item The interaction between Vagueness and Instruction format changes sign, and is now non-significant (beta=-.022, p=.374). 
\end{description}

@@

This experiment tested whether vague instructions would result in faster responses than crisp instructions, when borderline cases were present. Faster responses for vague instructions were found in pilot experiment B, but there were no borderline cases in that experiment.

In this experiment we found in contrast that vague instructions resulted in slower responses than crisp instructions: a difference that was significant when considering the full data (112ms), but which was not significant after removing the smallest arrays from the analysis, which had a pattern opposite to the main trends in the rest of the data.

We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average, such that the disadvantage of numerical format overwhelmed the contribution of vagueness. The verbal vague condition still yielded faster responses than the numerical crisp condition, so the pattern from pilot experiment B was reproduced, but in the light of the evidence from this experiment (Experiment 1), in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like an advantage of verbal instruction format.

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the difference between numeric and verbal) went hand in hand with might be called the (human) "selection algorithm": To see this, consider the task of selecting the dot array that contains "few dots":" to do this, it suffices to \emph{compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains "16 dots" seems to require the participant to estimate, and then \emph{match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use \emph{comparison} rather than having to resort to \emph{matching}.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. In Experiment 2 we restricted all the instructions to numeric quantifiers while factorially manipulating vagueness and selection task. In Experiment 3 we ensured that all instructions used verbal quantifiers, while also factorially manipulating vagueness and selection task. This allowed us to distinguish between the predictions of the selection task account and the instruction format account.
